# Steady State Distributions {.unnumbered}

{{< include _r_mc_setup.qmd >}}



## Markov's letters



```{r}

state_names = c("vowel", "consonant")

P = rbind(
  c(0.128, 0.872),
  c(0.663, 0.337)
)

pi_0 = c(0.432, 0.568)

```

### 1-step transition matrix

```{r}
plot_transition_matrix(P, state_names)
```

### 2-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 2)
```

### 3-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 3)
```


### 5-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 5)
```


### 10-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 10)
```


### 20-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 20)
```

## Ehrenfest urn chain


```{r}
M = 3

state_names = 0:M

P = rbind(c(0, 1, 0, 0),
         c(1/3, 0, 2/3, 0),
         c(0, 2/3, 0, 1/3),
         c(0, 0, 1, 0)
)

```

### 100-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 100)
```

### 101-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 101)
```


### Stationary distribution

```{r}
pi_s = compute_stationary_distribution(P)

# display in table
data.frame(state_names, t(pi_s)) |>
  kbl(col.names = c("state", "stationary probability")) |>
  kable_styling()

```

## Ping pong

```{r}
state_names = c("AB", "AC", "BC")

P = rbind(c(0, .7, .3),
          c(.8, 0, .2),
          c(.6, .4, 0)
)
```


### Stationry distribution

```{r}
pi_s = compute_stationary_distribution(P)

# display in table
data.frame(state_names, t(pi_s)) |>
  kbl(col.names = c("state", "stationary probability"), digits = 4) |>
  kable_styling()

```


### transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 1)
```

### 2-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 2)
```

### 10-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 10)
```


### 20-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 20)
```