# Markov Chains: Joint, Conditional, and Marginal Distributions {.unnumbered}

{{< include _r_mc_setup.qmd >}}


## Popcorn and ice cream


```{r}

state_names = c("ice cream", "popcorn")

P = rbind(
  c(0.2, 0.8),
  c(0.4, 0.6)
)

```


### Transition matrix

```{r}
plot_transition_matrix(P, state_names)

```

### 2-step transition matrix


```{r}
P %*% P
```


```{r}
library(expm)
P %^% 2
```



```{r}
plot_transition_matrix(P, n_step = 2)

```


### 3-step transition matrix

```{r}
P %^% 3
```


```{r}
plot_transition_matrix(P, n_step = 3)
```

## Weather chain


```{r}
state_names = c("RR", "NR", "RN", "NN")

P = rbind(c(.7, 0, .3, 0),
          c(.5, 0, .5, 0),
          c(0, .4, 0, .6),
          c(0, .2, 0, .8)
)
```


### Transition matrix

```{r}
plot_transition_matrix(P, state_names)

```

### 2-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 2)

```

### 2-step transition matrix

```{r}
plot_transition_matrix(P, state_names, n_step = 3)

```


## Ping pong

```{r}
state_names = c("AB", "AC", "BC")

P = rbind(c(0, .7, .3),
          c(.8, 0, .2),
          c(.6, .4, 0)
)
```

### Initial players chosen at random

```{r}
pi_0 = c(1/3, 1/3, 1/3)

pi_0 %*% P
```


```{r}
plot_DTMC_marginal_bars(pi_0, P, state_names, last_time = 10)
```


### Player A's initial opponent chosen at random

```{r}
pi_0 = c(1/2, 1/2, 0)

pi_0 %*% P
```


```{r}
plot_DTMC_marginal_bars(pi_0, P, state_names, last_time = 10)
```

### Player A and B play initially

```{r}
pi_0 = c(1, 0, 0)

pi_0 %*% P
```


```{r}
plot_DTMC_marginal_bars(pi_0, P, state_names, last_time = 10)
```




## Ehrenfest urn chain


```{r}
M = 3

state_names = 0:M

P = rbind(c(0, 1, 0, 0),
         c(1/3, 0, 2/3, 0),
         c(0, 2/3, 0, 1/3),
         c(0, 0, 1, 0)
)

```


### Molecules initially distributed at random between A and B


Marginal distribution of $X_0$

```{r}
pi_0 = dbinom(0:M, M, 0.5)

pi_0
```


Marginal distribution of $X_1$

```{r}
pi_0 %*% P
```


Marginal distribution of $X_2$

```{r}
pi_0 %*% (P %^% 2)
```

Marginal distribution of $X_3$

```{r}
pi_0 %*% (P %^% 3)
```



```{r}
plot_DTMC_marginal_bars(pi_0, P, state_names, last_time = 10)
```

### Molecules initially all in A

Marginal distribution of $X_0$

```{r}
pi_0 = c(rep(0, M), 1)

pi_0
```


Marginal distribution of $X_1$

```{r}
pi_0 %*% P
```


Marginal distribution of $X_2$

```{r}
pi_0 %*% (P %^% 2)
```

Marginal distribution of $X_3$

```{r}
pi_0 %*% (P %^% 3)
```



```{r}
plot_DTMC_marginal_bars(pi_0, P, state_names, last_time = 10)
```



### Molecules initially all in B

Marginal distribution of $X_0$

```{r}
pi_0 = c(1, rep(0, M))

pi_0
```


Marginal distribution of $X_1$

```{r}
pi_0 %*% P
```


Marginal distribution of $X_2$

```{r}
pi_0 %*% (P %^% 2)
```

Marginal distribution of $X_3$

```{r}
pi_0 %*% (P %^% 3)
```



```{r}
plot_DTMC_marginal_bars(pi_0, P, state_names, last_time = 10)
```

