# Discrete Time Markov Chains: Steady State Distributions {.unnumbered}

Suppose that bases (letters) in DNA sequences can be modeled as a Markov chain with state space (A, C, G, T) and transition matrix

```{r}
P = rbind(
  c(0.36, 0.15, 0.20, 0.29),
  c(0.41, 0.16, 0.04, 0.39),
  c(0.30, 0.21, 0.14, 0.35),
  c(0.31, 0.19, 0.18, 0.32)
)
```


```{r}
#| echo: false

library(kableExtra)

state_names = c("A", "C", "G", "T")

row.names(P) = state_names

kbl(P, col.names = state_names) |>
  column_spec(1, bold = TRUE) |>
  kable_styling()
```


(@) One of the following is the  unique stationary distribution.
Identify which one it is and explain your reasoning conceptually without doing any calculations.

\begin{align*}
\boldsymbol{\pi}_1 & = [0.1742, 0.3430, 0.3266, 0.1562]\\
\boldsymbol{\pi}_2 & = [0.3430, 0.1742, 0.1562, 0.3266]\\
\boldsymbol{\pi}_3 & = [0.4935, 0.0093, 0.0057, 0.4915]
\end{align*}


(@) Now use software to find the stationary distribution.

(@) Find the probability that C is followed two letters later by T.

(@) Find the probability that A is followed two letters later by T.

(@) Find the probability that C is followed 100 letters later by T.

(@) Find the probability that A is followed 100 letters later by T.

(@) Find the probability that a three letter sequences spell "CAT".
